Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-07-13 analysis/paper/paper.qmd
editing methods section

In our typological study we identified some faceted sub-spheroids/spheroids in the Qianshangying assemblage. To minimize subjective judgments that are inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to our 3d models to evaluate the degree of difference between the archaeological spheroids and a perfect sphere. In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.

These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact. We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies. Furthermore, we compared the results of the typological study and our new morphometric analyses to investigate the validity of the traditional typological approach.

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used the closed-source software Artefact3-D to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. @Noshita2025 . Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

@@

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].All these were conducted in python.

This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores. Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP. Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

@@

Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree. This provides a rotation-invariant measure of shape complexity across spatial frequencies. For normalized shapes, this total energy approaches 1 for a perfect sphere, while more irregular forms yield higher values.

@@

We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023], which was computed in python. A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset. As the object becomes more irregular or asymmetric, this offset increases. For each 3D model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box. The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result. The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms.

@@

To investigate the production strategies of spheroids in QSY, this part aims to determine whether these artifacts were intentionally shaped and whether their manufacture followed standardized procedures. Addressing this question is crucial for evaluating whether spheroids represent a distinct technological category, separate from exhausted cores.

@@

Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids—that is, the removal patterns and associated stages employed in a spheroid. Ideally, the goal is to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product. If consistent patterns are observed, they may indicate shared reduction schemes or conceptual templates. Furthermore, we explore whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.

>>>>

In our typological study we identified some faceted sub-spheroids/spheroids in the Qianshangying assemblage. To minimize subjective judgments that are inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to our 3d models to evaluate the degree of difference between the archaeological spheroids and a perfect sphere. In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids. These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact. We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies. Furthermore, we compared the results of the typological study and our new morphometric analyses to investigate the validity of the traditional typological approach.

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. They find that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items, however they did not provide code to document their analysis. @Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250). Their results were consistent with the  hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced. Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM.  We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

@@

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023]. We included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP as controls to aid in the interpretation of the results. Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics. All these were conducted in Python. This study represents the first attempt that we are aware of to apply an open source and reproducible SPHARM analysis for the classification of stone artefacts.

@@

Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree. This provides a rotation-invariant measure of shape complexity across spatial frequencies. For normalized shapes, this total energy approaches one for a perfect sphere, while more irregular forms yield higher values.

@@

We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023]. A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset. As the object becomes more irregular or asymmetric, this offset increases. For each 3D model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box. The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result. The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms.

@@

To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying. Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures. Addressing this question is crucial for achieving our goal of determining whether or not spheroids represent a distinct technological category, separate from exhausted cores.

@@

Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids—that is, the removal patterns and associated stages employed in a spheroid. Our goal was to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product. If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.We also explored whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.
